First Order Alternation

### 2018-09-11 quantifiers/draft.tex
ok

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. To simplify the
presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
is an infinite domain and $\Bool$ is the set of truth values $\top$
(true) and $\bot$ (false). The $\Data$ sort is equipped with finitely
many function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
$\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
$\#(f)=0$, we say that $f$ is a constant. A \emph{predicate} is a
function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
relation of arity $\#(p)$ and we write $\preds$ for the set of
predicates. In the following, we shall consider that the
interpretation of all function symbols $f : \Data^{\#(f)} \rightarrow
\Data$ that are not predicates is fixed and relative to the
interpretation of the $\Data$ sort. For simplicity, we further blur
the notational distinction between function symbols and their
interpretations.

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over the interpretation of the $\Data$ sort. Terms
are either constants of sort $\Data$, variables or function
applications. The set of first-order formulae is defined by the syntax
below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. A \emph{sentence}
is a formula $\phi$ in which each variable occurs under the scope of a
quantifier, i.e. $\fv{}{\phi} = \emptyset$. A formula is
\emph{positive} if each predicate symbol occurs under an even number
of negations. The size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
For a formula $\phi$, we define $\sem{\phi}_\nu \isdef \set{\I \mid
  \I,\nu \models \phi}$, where $\nu(x)$ is defined for each $x \in
\fv{}{\phi}$, and drop the subscript for sentences. A sentence $\phi$
is \emph{satisfiable} (\emph{unsatisfiable}) if
$\sem{\phi}\neq\emptyset$ [$\sem{\phi}=\emptyset$]. An element of
$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
and every valuation $\nu$.

@@

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X$, we denote by $X
\mapsto \Data$ the set of valuations of the variables $X$ and
$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
set of \emph{data symbols} $(a,\nu)$, where $a$ is an input symbol and
$\nu$ is a valuation. A \emph{data word} (simply called word in the
following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

A \emph{first-order alternating automaton} (simply called automaton in
the following) is a tuple $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$,
where $\Sigma$ is the finite event alphabet, $X$ is a finite set of
variables, $Q$ is a finite set of predicates, $\iota$ is a satisfiable
sentence, $F \subseteq Q$ is the set of \emph{final} predicates and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event, $X \cap
\set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive formula
such that $\fv{}{\phi} \subseteq X \cup \set{y_1,\ldots,y_{\#(q)}}$
and each predicate that occurs in $\psi$ belongs also to $Q$. The
\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I, q \in Q}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$. Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
\Sigma[X]^*$, an \emph{execution} of
$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each tree $T_i$ is
such that:
\begin{compactitem}
\item the roots of $T_1, T_2, \ldots$ form a cube from
  $\cube{\minsem{\iota}}$, % --- because $\iota$ is satisfiable, such
                           % a cube always exists,
%
% \item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
  [n-1]$ in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$. % --- because $\fv{}{\psi} \subseteq X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$, the values assigned
                  % by $\eta$ to variables outside $X \cup
                  % \set{y_1,\ldots,y_{\#(q)}}$ have no impact on
                  % $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
An execution $\T$ over $w$ is said to be \emph{accepting} if and only
if \begin{inparaenum}[(i)]
\item all paths in $\T$ have the same length $n$, and
\item the frontier of each tree $T \in \T$ is labeled with \emph{final
  configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$ is a
  final predicate.
\end{inparaenum} In this case, we say that $\A$ accepts $w$ and
denote by $\lang{\A}$ the set of words accepted by $\A$. We address
the following questions: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

In the upcoming developments it is more convenient to work with
logical formulae defining executions. Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. As a shorthand, we write $\stamp{Q}{\leq n}$ (resp.
$\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid q \in Q, i
  \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in [n]}$). For
a formula $\psi$ and $i \in \nat$, we define $\stamp{\psi}{i} \isdef
\psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the formula in which all input
variables and state predicates (and only those symbols) are replaced
by their time-stamped counterparts. As a shorthand, we shall write
$q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.

Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
\emph{path formula} of $\alpha$ (in $\A$) is defined as follows:
\begin{equation}\label{eq:pathform}
\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
\end{equation}
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\T$ of $\A$ over $w$ is associated an interpretation
$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
defined as:
\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
  q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\T$}}\]

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\T} \mid \T
    \text{ is an execution of $\A$ over $w$}}\). 
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be
  the subformula of $\pathform{w_\Sigma}$ corresponding to the
  application(s) of the transition rule $q(\vec{y})
  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
  [1,\#(q)]$ and behaves like $w_\Data$ outside of
  $\set{y_1,\ldots,y_{\#(d)}}$. Because $\I, w_\Data \models \forall
  y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$,
  we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
  the minimal interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

>>>>

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. We consider two
sorts $\Data$ and $\Bool$, where $\Data$ is an infinite domain and
$\Bool = \set{\top,\bot}$ is the set of boolean values true ($\top$)
and false ($\bot$), respectively. The $\Data$ sort is equipped with
finitely many function symbols $f : \Data^{\#(f)} \rightarrow \Data$,
where $\#(f)\geq0$ denotes the number of arguments (arity) of
$f$. When $\#(f)=0$, we say that $f$ is a constant. A \emph{predicate}
is a function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
relation of arity $\#(p)$ and we write $\preds$ for the set of
predicates.

In the following, we shall consider that the interpretation of all
function symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not
predicates is fixed by the interpretation of the $\Data$ sort,
e.g.\ if $\Data$ is the set of integers $\zed$, the function symbols
are zero, the successor function and the arithmetic operations of
addition and multiplication. For simplicity, we further blur the
notational distinction between function symbols and their
interpretations.

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
formulae is defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. A \emph{sentence}
is a formula $\phi$ in which each variable occurs under the scope of a
quantifier, i.e. $\fv{}{\phi} = \emptyset$. A formula is
\emph{positive} if each predicate symbol occurs under an even number
of negations. The size $\len{\phi}$ of a formula $\phi$ is the number
of symbols needed to write it down.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
is \emph{satisfiable} (\emph{unsatisfiable}) if
$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
and every valuation $\nu$.

@@

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X \subseteq \vars$, we denote by $X
\mapsto \Data$ the set of valuations of the variables $X$ and
$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
set of \emph{data symbols} $(a,\nu)$, where $a$ is an input symbol and
$\nu$ is a valuation. A \emph{data word} (simply called word in the
following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

A \emph{first-order alternating automaton} (automaton) is a tuple $\A
= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
\item $\Sigma$ is the finite set of input events, 
\item $X$ is a finite set of input variables, 
\item $Q$ is a finite set of predicates denoting control states, 
\item $\iota$ is a sentence defining initial configurations,
\item $F \subseteq Q$ is the set of predicates denoting final states, and 
\item $\Delta$ is a set of \emph{transition rules} of the form
  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
  a predicate, $a \in \Sigma$ is an input event, $X \cap
  \set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive
  formula such that $\fv{}{\phi} \subseteq X \cup
  \set{y_1,\ldots,y_{\#(q)}}$ and each predicate that occurs (under an
  even number of negations) in $\psi$ belongs also to $Q$. The
  quantifiers occurring in $\psi$ are referred to as \emph{transition
    quantifiers}.
\end{compactitem}
The \emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$.

Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
$T_i$ is a tree labeled with configurations, such that:
\begin{compactitem}
\item the labels of the roots of $T_1, T_2, \ldots$ form a cube from
  $\cube{\minsem{\iota}}$,
%
\item if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
  [n-1]$ in $T_i$, then the labels of its children form a cube from
  $\cube{\minsem{\psi}_{\nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)}
        \leftarrow d_{\#(q)}]}}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule.
\end{compactitem}
An execution $\T$ over $w$ is said to be \emph{accepting} if and only
if \begin{inparaenum}[(i)]
\item all paths in $\T$ have the same length $n$, and
\item the frontier of each tree $T \in \T$ is labeled with \emph{final
  configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$.
\end{inparaenum} In this case, we say that $\A$ accepts $w$ and
denote by $\lang{\A}$ the set of words accepted by $\A$.

In this paper, we address the following questions: \begin{compactenum}
\item \emph{boolean closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

In the upcoming developments it is sometimes more convenient to work
with logical formulae defining executions of automata, than with
low-level execution forests. For this reason, we first introduce
\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
the executions of an automaton, over words that share a given sequence
$\alpha$ of input events.  Second, we restrict a path formula
$\pathform{\alpha}$ to an \emph{acceptance formula}
$\accform{\alpha}$, which defines only accepting executions over words
that share a given input sequence. Otherwise stated,
$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
a word $w$ such that $w_\Sigma = \alpha$.

Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicates
and variables, respectively. As a shorthand, we write $\stamp{Q}{\leq
  n}$ (resp.  $\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid
  q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in
  [n]}$). For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates (and only
those symbols) are replaced by their time-stamped counterparts. As a
shorthand, we shall write $q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$,
when no confusion arises.

Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
\emph{path formula} of $\alpha$ is defined as:
\begin{equation}\label{eq:pathform}
\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
\end{equation}
The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
be clear from the context. To formalize the relation between the
low-level configuration-based execution semantics and the symbolic
path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
execution forest $\T$ of $\A$ over $w$ is associated an interpretation
$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
defined as:
\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}}\]

\begin{lemma}\label{lemma:path-formula}
  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\T} \mid \T
    \text{ is an execution of $\A$ over $w$}}\). 
\end{lemma}
\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
  minimal. By the induction hypothesis, there exists an execution $\G$
  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be the
  subformula of $\pathform{w_\Sigma}$ corresponding to the
  application(s) of the transition rule $q(\vec{y}) \arrow{a_n}{}
  \psi$ at the $(n-1)$-th step. Let $\nu = w_\Data[y_1 \leftarrow
    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, w_\Data
  \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$, we have $\I
  \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of the minimal
  interpretations such that $\K \subseteq \I$ and $\K \in
  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
  and is unique, otherwise we could take the pointwise intersection of
  two or more such interpretations. We define the interpretation
  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
  contradiction. Then we extend the execution $\G$ by appending to
  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
  the cube $\cube{\overline{\K}}$. By repeating this step for all
  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.

  ``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
  that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
  w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
  0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
  definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
  is a minimal such interpretation. For the inductive case $n > 0$,
  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
  restriction of $\T$ to $u$. Consequently, $\I_\G$ is the restriction
  of $\I_\T$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
  \models \pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
    \text{ labels a node on the $n$-th level in $\T$}}$, we have $\I_{\T},
  w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y})
  \rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
  execution semantics of $\A$. This is the case because the children
  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
  $(n-1)$-th level of $\T$ form a cube from
  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
  each $y_i$ the value $d_i$ and behaves like $w_\Data$,
  otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
  minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
  that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
  must coincide with $\I_{\G}$. Assuming this is not the case,
  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
  \subsetneq \I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
  transition rules in which the predicate symbol $q$ occurs on the
  right-hand side. Then it must be the case that, for some node on the
  $(n-1)$-th level of $\G$, labeled with a configuration
  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
  contradicts the execution semantics of $\A$. \qed}
